---
title: "Tracking the Success of Nations"
author: "Chace Paulson"
date: 12-12-2018
output: pdf_document
---

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# Sovereign Nations -------------------------------------------------------

# import sovereign nations data set 
library(readxl)
sovereign_nations <- read_excel("~/Downloads/final_proj/raw_data/sovereign_nations.xlsx")

library(tidyverse)
sovereign_nations <- as_tibble(sovereign_nations)

# rename columns
library(magrittr)
sovereign_nations <- sovereign_nations %>% 
  rename(country_name = `Soverign Nations`)

# Political Freedom -------------------------------------------------------

# load political freedom data 
library(readxl)
poli_freedom <- read_excel("~/Downloads/final_proj/raw_data/poli_freedom.xlsx")

library(tidyverse)
poli_freedom <- as_tibble(poli_freedom)

# rename columns in political freedom data 
colnames(poli_freedom)
poli_freedom <- poli_freedom %>% 
  rename(country_name = `Country/Territory`, 
         poli_status = Status, 
         poli_rights_rating = `PR Rating`, 
         civil_lib_rating = `CL Rating`, 
         poli_freedom = Total
  )

# create second data set only with columns we care about 
small_poli <- tibble("country_name" = poli_freedom$country_name, 
                         "poli_freedom" = poli_freedom$poli_freedom)

# Economic Freedom --------------------------------------------------------

# Load economic freedom data 
library(readxl)
econ_freedom <- read_excel("~/Downloads/final_proj/raw_data/econ_freedom.xls")

library(tidyverse)
econ_freedom <- as_tibble(econ_freedom)
colnames(econ_freedom)

# Rename columns in economic freedom data 
econ_freedom <- econ_freedom %>% 
  rename(
    countryID = `CountryID`, 
    country_name = `Country Name`,
    webname = WEBNAME, 
    region = Region, 
    world_rank = `World Rank`, 
    region_rank = `Region Rank`, 
    total_score = `2018 Score`, 
    property_rights = `Property Rights`, 
    judical_effectiveness = `Judical Effectiveness`, 
    govt_integrity = `Government Integrity`, 
    tax_burden = `Tax Burden`, 
    govt_spending = `Gov't Spending`, 
    fiscal_health = `Fiscal Health`, 
    bis_freedom = `Business Freedom`, 
    labor_freedom = `Labor Freedom`, 
    monetary_freedom = `Monetary Freedom`, 
    trade_freedom = `Trade Freedom`, 
    invest_freedom = `Investment Freedom`, 
    finance_freedom = `Financial Freedom`, 
    tariff_rate_perc = `Tariff Rate (%)`, 
    income_tax_rate_per = `Income Tax Rate (%)`, 
    corporate_tax_rate_per = `Corporate Tax Rate (%)`, 
    tax_burden_per_of_gdp = `Tax Burden % of GDP`, 
    govt_expenditure_per_of_gdp = `Gov't Expenditure % of GDP`, 
    country = Country, 
    pop_in_mil = `Population (Millions)`, 
    gdp_bill_ppp = `GDP (Billions, PPP)`, 
    gdp_growth_rate_perc = `GDP Growth Rate (%)`, 
    five_year_gdp_growth_rate_per = `5 Year GDP Growth Rate (%)`, 
    gdp_per_capita = `GDP per Capita (PPP)`, 
    unemployment_per = `Unemployment (%)`, 
    inflation_per = `Inflation (%)`, 
    fdi_inflow_mil = `FDI Inflow (Millions)`, 
    public_debt_per_of_gdp = `Public Debt (% of GDP)`
  )

colnames(econ_freedom)

# create second data set only with columns we care about 

small_econ <- tibble("country_name" = econ_freedom$country_name,
                         "region" = econ_freedom$region,
                         "econ_freedom" = econ_freedom$total_score)

# GDP ---------------------------------------------------------------------

# loading gdp data 
library(readxl)
gdp <- read_excel("~/Downloads/final_proj/raw_data/gdp.xlsx")

library(tidyverse)
gdp <- as_tibble(gdp)

# renaming columns in gdp data 
colnames(gdp)
gdp <- gdp %>% 
  rename(country_name = `Country Name`, 
         country_code = `Country Code`, 
         indicator_name = `Indicator Name`, 
         indicator_code = `Indicator Code`)
colnames(gdp)

# finding out if there is any missing data from the 2017 set 
sum(is.na(gdp$`2017`))

# merge with sovereign nations to keep only desired nations
gdp_mixed <- merge(sovereign_nations, gdp, all.x = TRUE)

# find out if there is still missing data from 2017 set now that we only 
# have the nations we are interested in 
sum(is.na(gdp_mixed$`2017`))
gdp_mixed$country_name[is.na(gdp_mixed$`2017`)]

# create a column of gdp_total to combine other years in to 
gdp_mixed$gdp_total <- gdp_mixed$`2017`

# for the countries with NAs for 2017, we will take data from 2016 instead
gdp_mixed$gdp_total[is.na(gdp_mixed$gdp_total)] <- 
  as.character(gdp_mixed$`2016`[is.na(gdp_mixed$gdp_total)]) 

# test for number of remaining NAs
sum(is.na(gdp_mixed$gdp_total))

# repeat but using 2015 data 
gdp_mixed$gdp_total[is.na(gdp_mixed$`gdp_total`)] <- 
  as.character(gdp_mixed$`2015`[is.na(gdp_mixed$gdp_total)]) 
sum(is.na(gdp_mixed$gdp_total))

# repeat using 2014 data 
gdp_mixed$gdp_total[is.na(gdp_mixed$`gdp_total`)] <- 
  as.character(gdp_mixed$`2014`[is.na(gdp_mixed$gdp_total)]) 
sum(is.na(gdp_mixed$gdp_total))

# repeat using 2013 data 
gdp_mixed$gdp_total[is.na(gdp_mixed$`gdp_total`)] <- 
  as.character(gdp_mixed$`2013`[is.na(gdp_mixed$gdp_total)]) 
sum(is.na(gdp_mixed$gdp_total))

# repeat using 2013 data 
gdp_mixed$gdp_total[is.na(gdp_mixed$`gdp_total`)] <- 
  as.character(gdp_mixed$`2012`[is.na(gdp_mixed$gdp_total)]) 
sum(is.na(gdp_mixed$gdp_total))

# since we continue to having a total of 6 NAs, we will stop here

# find out what countries are still missing gdp data
gdp_mixed$country_name[is.na(gdp_mixed$gdp_total)]

# create smaller data set with only the gdp_total and country_names columns
small_gdp <- tibble("country_name" = gdp_mixed$country_name, 
                        "gdp" = gdp_mixed$gdp_total)

# Unemployment Rate -------------------------------------------------------

# import unemployment data 
library(readxl)
unemployment <- read_excel("~/Downloads/final_proj/raw_data/unemployment.xlsx")

# renaming columns 
unemployment <- unemployment %>% 
  rename(country_name = `Country`, 
         unemployment_rate = `%`, 
         date = `Date of Info`)

# create small data set with just country_name and unemployment rate
small_unemployment <- tibble("country_name" = unemployment$country_name, 
                                 "unemployment" = unemployment$unemployment_rate)

# Crime Index -------------------------------------------------------------

# loading crime index data 
library(readxl)
crime_index <- read_excel("~/Downloads/final_proj/raw_data/crime_index.xlsx")

library(tidyverse)
crime_index <- as_tibble(crime_index)

# renaming columns in crime index data 
colnames(crime_index)
crime_index <- crime_index %>% 
  rename(rank = `Rank`,
         country_name = `Country Name`, 
         crime_index = `Crime Index`, 
         safety_index = `Safety Index`)
colnames(crime_index)

# merge with sovereign nations to keep only desired nations
crime_mixed <- merge(sovereign_nations, crime_index, all.x = TRUE)

# test for NAs 
sum(is.na(crime_mixed$crime_index))
crime_mixed$country_name[is.na(crime_mixed$crime_index)]

# create smaller tibble keeping only country_name and literacy
small_crime <- tibble("country_name" = crime_mixed$country_name, 
                          "crime_index" = crime_mixed$crime_index)

# Homicide Rate -----------------------------------------------------------

# loading homicide data 
library(readxl)
homicide <- read_excel("~/Downloads/final_proj/raw_data/homicide_rates.xlsx")

library(tidyverse)
homicide <- as_tibble(homicide)

# renaming columns in homicide data 
colnames(homicide)
homicide <- homicide %>% 
  rename(country_name = `Country Name`)
colnames(homicide)

# finding out if there is any missing data from the 2016 set 
sum(is.na(homicide$`2016`))

# merge with sovereign nations to keep only desired nations
homicide_mixed <- merge(sovereign_nations, homicide, all.x = TRUE)

# find out if there is still missing data from 2016 set now that we only 
# have the nations we are interested in 
sum(is.na(homicide_mixed$`2016`))
homicide_mixed$country_name[is.na(homicide_mixed$`2016`)]

# create a column of median_total to combine other years in to 
homicide_mixed$homicide_total <- homicide_mixed$`2016`

# for the countries with NAs for 2016, we will take data from 2015 instead
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2015`[is.na(homicide_mixed$homicide_total)]) 

# test for number of remaining NAs
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2014 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2014`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2013 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2013`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2012 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2012`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2011 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2011`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2010 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2010`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2009 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2009`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2008 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2008`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2007 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2007`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2006 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2006`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2005 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2005`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# repeat but using 2004 data
homicide_mixed$homicide_total[is.na(homicide_mixed$homicide_total)] <- 
  as.character(homicide_mixed$`2004`[is.na(homicide_mixed$homicide_total)]) 
sum(is.na(homicide_mixed$homicide_total))

# since we continue to having a total of 3 NAs, we will stop here

# find out what countries are still missing homicide data
homicide_mixed$country_name[is.na(homicide_mixed$homicide_total)]

# create smaller data set with only the homicide_total and country_names columns
small_homicide <- tibble("country_name" = homicide_mixed$country_name, 
                             "homicide" = homicide_mixed$homicide_total)

# Secondary School Enrollment Rate ----------------------------------------

# loading enrollment data 
library(readxl)
enrollment <- read_excel("~/Downloads/final_proj/raw_data/secondary_enrollment.xlsx")

library(tidyverse)
enrollment <- as_tibble(enrollment)

# renaming columns in enrollment data 
colnames(enrollment)
enrollment <- enrollment %>% 
  rename(country_name = `Country Name`)
colnames(enrollment)

# finding out if there is any missing data from the 2016 set 
sum(is.na(enrollment$`2016`))

# merge with sovereign nations to keep only desired nations
enrollment_mixed <- merge(sovereign_nations, enrollment, all.x = TRUE)

# find out if there is still missing data from 2016 set now that we only 
# have the nations we are interested in 
sum(is.na(enrollment_mixed$`2016`))
enrollment_mixed$country_name[is.na(enrollment_mixed$`2016`)]

# create a column of median_total to combine other years in to 
enrollment_mixed$enrollment_total <- enrollment_mixed$`2016`

# for the countries with NAs for 2016, we will take data from 2015 instead
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2015`[is.na(enrollment_mixed$enrollment_total)]) 

# test for number of remaining NAs
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2014 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2014`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2013 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2013`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2012 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2012`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2011 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2011`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2010 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2010`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2009 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2009`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2008 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2008`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2007 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2007`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2006 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2006`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2005 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2005`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2004 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2004`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2003 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2003`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2002 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2002`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2001 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2001`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# repeat but using 2000 data
enrollment_mixed$enrollment_total[is.na(enrollment_mixed$enrollment_total)] <- 
  as.character(enrollment_mixed$`2000`[is.na(enrollment_mixed$enrollment_total)]) 
sum(is.na(enrollment_mixed$enrollment_total))

# since we continue to having a total of 11 NAs, we will stop here

# merge with sovereign nations to keep only desired nations
enrollment_mixed <- merge(sovereign_nations, enrollment_mixed, all.x = TRUE)

# find out what countries are still missing enrollment data
sum(is.na(enrollment_mixed$enrollment_total))
enrollment_mixed$country_name[is.na(enrollment_mixed$enrollment_total)]

# create smaller data set with only the enrollment_total and country_names columns
small_enrollment <- tibble("country_name" = enrollment_mixed$country_name, 
                               "enrollment" = enrollment_mixed$enrollment_total)

# Literacy Rate -----------------------------------------------------------

# import literacy rates data set 
library(readxl)
literacy <- read_excel("~/Downloads/final_proj/raw_data/literacy_rates2.xlsx")

library(tidyverse)
literacy <- as_tibble(literacy)

# rename columns in the data set
library(magrittr)
colnames(literacy)
literacy <- literacy %>% 
  rename(country_name = `COUNTRY`, 
         definition = `DEFINITION`, 
         total = `TOTAL POPULATION`, 
         male = `MALE`, 
         female = `FEMALE`, 
         note = `NOTE`)
colnames(literacy)
head(literacy)

# split the strings to start removing repetition
literacy2 <- str_split_fixed(literacy$definition, ": ", n = 2)
literacy3 <- str_split_fixed(literacy$total, ": ", n = 2)
literacy4 <- str_split_fixed(literacy$male, ": ", n = 2)
literacy5 <- str_split_fixed(literacy$female, ": ", n = 2)

# convert to tibbles
literacy2 <- as.tibble(literacy2)
literacy3 <- as.tibble(literacy3)
literacy4 <- as.tibble(literacy4)
literacy5 <- as.tibble(literacy5)

# rename column names
literacy2 %<>% 
  rename(definition = `V2`)
literacy3 %<>% 
  rename(total = `V2`)
literacy4 %<>% 
  rename(male = `V2`)
literacy5 %<>% 
  rename(female = `V2`)

# create new data frame with only the columns of interest
fixed_literacy <- tibble("country_name" = literacy$country_name, 
                             "definition" = literacy2$definition, 
                             "total" = literacy3$total, 
                             "male" = literacy4$male, 
                             "female" = literacy5$female, 
                             "note" = literacy$note)
fixed_literacy <- as.tibble(fixed_literacy)

# remove percent sign from data 
fixed_literacy$total <- 
  substr(fixed_literacy$total, 1, 
         nchar(fixed_literacy$total)-1)

# merge with sovereign nations to keep only desired nations
literacy_mixed <- merge(sovereign_nations, fixed_literacy, all.x = TRUE)

# test for NAs 
sum(is.na(literacy_mixed$total))
literacy_mixed$country_name[is.na(literacy_mixed$total)]

# create smaller tibble keeping only country_name and literacy
small_literacy2 <- tibble("country_name" = literacy_mixed$country_name, 
                              "literacy" = literacy_mixed$total)

# Life Expectancy Rate ----------------------------------------------------

# import life expectancy rates data set 
library(readxl)
life_expectancy <- read_excel("~/Downloads/final_proj/raw_data/life_expectancy.xlsx")

library(tidyverse)
life_expectancy <- as_tibble(life_expectancy)

# rename columns in the data set
library(magrittr)
colnames(life_expectancy)
life_expectancy <- life_expectancy %>% 
  rename(country_name = `Country Name`, 
         life_expectancy = `Life Expectancy`)
colnames(life_expectancy)

# split the string in column 1 to put the country_name in the second column
life2 <- NA
life2 <- str_split_fixed(life_expectancy$country_name, " ", n = 2)

# convert to tibble and rename columns
life2 <- as.tibble(life2)
life2 <- life2 %>% 
  rename(country_id = `V1`, 
         country_name = `V2`) 
life2$life_expectancy <- NA

# trim the leading white spaces off of the country_name column
head(life2)
life2$country_name <- trimws(life2$country_name)

# put the life_expectancy rates into third column
library(dplyr)
life2 <- life2 %>% 
  separate(country_name, into = c("country_name", "life_expectancy"), 
           sep = "(?<=[a-zA-Z])\\s*(?=[0-9])")
head(life2)

# merge with sovereign nations to keep only desired nations
life_mixed <- merge(sovereign_nations, life2, all.x = TRUE)

# test for NAs 
sum(is.na(life_mixed$life_expectancy))
life_mixed$country_name[is.na(life_mixed$life_expectancy)]

# create smaller tibble keeping only country_name and life_expectancy 
small_life <- tibble("country_name" = life_mixed$country_name, 
                         "life_expectancy" = life_mixed$life_expectancy)

# Infant Mortality Rate ---------------------------------------------------

# import infant mortality rates data set 
library(readxl)
infant_mortality <- read_excel("~/Downloads/final_proj/raw_data/infant_mortality.xlsx")

library(tidyverse)
infant_mortality <- as_tibble(infant_mortality)

# rename columns in the data set
library(magrittr)
colnames(infant_mortality)
infant_mortality <- infant_mortality %>% 
  rename(country_name = `Country Name`, 
         infant_mortality = `Infant Mortality Rate`)
colnames(infant_mortality)

# split the string in column 1 to put the country_name in the second column
infant2 <- NA
infant2 <- str_split_fixed(infant_mortality$country_name, " ", n = 2)

# convert to tibble and rename columns
infant2 <- as.tibble(infant2)
infant2 <- infant2 %>% 
  rename(country_id = `V1`, 
         country_name = `V2`) 
infant2$infant_mortality <- NA

# trim the leading white spaces off of the country_name column
head(infant2)
infant2$country_name <- trimws(infant2$country_name)

# put the infant_mortality rates into third column
library(dplyr)
infant2 <- infant2 %>% 
  separate(country_name, into = c("country_name", "infant_mortality"), 
           sep = "(?<=[a-zA-Z])\\s*(?=[0-9])")
head(infant2)

# merge with sovereign nations to keep only desired nations
infant_mixed <- merge(sovereign_nations, infant2, all.x = TRUE)

# test for NAs 
sum(is.na(infant_mixed$infant_mortality))
infant_mixed$country_name[is.na(infant_mixed$infant_mortality)]

# create smaller tibble keeping only country_name and infant_mortality 
small_infant <- tibble("country_name" = infant_mixed$country_name, 
                           "infant_mortality" = infant_mixed$infant_mortality)

# Median Income -----------------------------------------------------------

# loading median income data 
library(readxl)
median_income <- read_excel("~/Downloads/final_proj/raw_data/median_income.xlsx")

library(tidyverse)
median_income <- as_tibble(median_income)

# renaming columns in median income data 
colnames(median_income)
median_income <- median_income %>% 
  rename(country_name = `Country Name`)
colnames(median_income)

# finding out if there is any missing data from the 2016 set 
sum(is.na(median_income$`2016`))

# merge with sovereign nations to keep only desired nations
median_mixed <- merge(sovereign_nations, median_income, all.x = TRUE)

# find out if there is still missing data from 2016 set now that we only 
# have the nations we are interested in 
sum(is.na(median_mixed$`2016`))
median_mixed$country_name[is.na(median_mixed$`2016`)]

# create a column of median_total to combine other years in to 
median_mixed$median_total <- median_mixed$`2016`

# for the countries with NAs for 2016, we will take data from 2015 instead
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2015`[is.na(median_mixed$median_total)]) 

# test for number of remaining NAs
sum(is.na(median_mixed$median_total))

# repeat but using 2014 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2014`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# repeat but using 2013 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2013`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# repeat but using 2012 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2012`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# repeat but using 2011 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2011`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# repeat but using 2010 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2010`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# repeat but using 2009 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2009`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# repeat but using 2008 data 
median_mixed$median_total[is.na(median_mixed$median_total)] <- 
  as.character(median_mixed$`2008`[is.na(median_mixed$median_total)]) 
sum(is.na(median_mixed$median_total))

# since we continue to having a total of 20 NAs, we will stop here

# find out what countries are still missing median income data
median_mixed$country_name[is.na(median_mixed$median_total)]

# create smaller data set with only the median_total and country_names columns
small_median_income <- tibble("country_name" = median_mixed$country_name, 
                                  "median_income" = median_mixed$median_total)

# World Happiness Rating --------------------------------------------------

# import happiness data set 
library(readxl)
happiness <- read_excel("~/Downloads/final_proj/raw_data/world_happiness_rating.xlsx")

library(tidyverse)
happiness <- as_tibble(happiness)

# rename columns in the data set
library(magrittr)
colnames(happiness)
happiness <- happiness %>% 
  rename(country_name = `Country Name`, 
         happiness_rating = `World Happiness Rating`)
colnames(happiness)

# split the string in column 1 to put the ratings in the second column
happiness <- happiness %>% 
  separate(country_name, into = c("country_name", "happiness_rating"), sep = "\\(")

# removing the white space at the end of the country name
happiness$country_name <- trimws(happiness$country_name)

# removing the parentheses at the end of the happiness rating
happiness$happiness_rating <- 
  substr(happiness$happiness_rating, 1, 
         nchar(happiness$happiness_rating)-1)

# merge with sovereign nations to keep only desired nations
happiness_mixed <- merge(sovereign_nations, happiness, all.x = TRUE)

# test for NAs 
sum(is.na(happiness_mixed))
happiness_mixed$country_name[is.na(happiness_mixed$happiness_rating)]

# Success Data ------------------------------------------------------------

# merge all of the small_econ and small_poli data into one data frame 
small_data <- merge(small_econ, small_poli, all.x = TRUE, all.y = TRUE)

# check if there are countrys not in our sovereign_nation data in 
# the small_data
setdiff(sovereign_nations$country_name, small_data$country_name)
setdiff(small_data$country_name, sovereign_nations$country_name)

# eliminate all countrys not in our sovereign_nation data 
success_data <- merge(sovereign_nations, small_data, all.x = TRUE)

# check that only the countrys from the soverign_nation data remain
nrow(success_data)
nrow(sovereign_nations)
setdiff(sovereign_nations$country_name, success_data$country_name)
setdiff(success_data$country_name, sovereign_nations$country_name)

# merge in the gdp data 
success_data <- merge(success_data, small_gdp, all.x = TRUE)

# merge in the unemployment data 
success_data <- merge(success_data, small_unemployment, all.x = TRUE)

# merge in crime rate data
success_data <- merge(success_data, small_crime, all.x = TRUE)

# merge in homicide data 
success_data <- merge(success_data, small_homicide, all.x = TRUE)

# merge in high school enrollment data
success_data <- merge(success_data, small_enrollment, all.x = TRUE)

# merge in the literacy data 
success_data <- merge(success_data, small_literacy2, all.x = TRUE)

# merge in the life expectancy data 
success_data <- merge(success_data, small_life, all.x = TRUE)

# merge in the infant mortality data 
success_data <- merge(success_data, small_infant, all.x = TRUE)

# merge in the median income data 
success_data <- merge(success_data, small_median_income, all.x = TRUE)

# merge in the happiness data 
success_data <- merge(success_data, happiness_mixed, all.x = TRUE)

# check to make sure all columns have been inputted
colnames(success_data)
ncol(success_data)

# turn data frame into tibble
success_data <- as.tibble(success_data)

# find the types for each column 
typeof(success_data$econ_freedom)
typeof(success_data$poli_freedom)
typeof(success_data$gdp)
typeof(success_data$unemployment)
typeof(success_data$crime_index)
typeof(success_data$homicide)
typeof(success_data$enrollment)
typeof(success_data$literacy)
typeof(success_data$life_expectancy)
typeof(success_data$infant_mortality)
typeof(success_data$median_income)
typeof(success_data$happiness_rating)

# convert the character strings into numeric 
success_data$econ_freedom <- 
  as.numeric(as.character(success_data$econ_freedom))
success_data$gdp <- 
  as.numeric(as.character(success_data$gdp))
success_data$homicide <- 
  as.numeric(as.character(success_data$homicide))
success_data$enrollment <- 
  as.numeric(as.character(success_data$enrollment))
success_data$literacy <- 
  as.numeric(as.character(success_data$literacy))
success_data$life_expectancy <- 
  as.numeric(as.character(success_data$life_expectancy))
success_data$infant_mortality <- 
  as.numeric(as.character(success_data$infant_mortality))
success_data$median_income <- 
  as.numeric(as.character(success_data$median_income))
success_data$happiness_rating <- 
  as.numeric(as.character(success_data$happiness_rating))

# check to ensure that all columns are now doubles
typeof(success_data$econ_freedom)
typeof(success_data$poli_freedom)
typeof(success_data$gdp)
typeof(success_data$unemployment)
typeof(success_data$crime_index)
typeof(success_data$homicide)
typeof(success_data$enrollment)
typeof(success_data$literacy)
typeof(success_data$life_expectancy)
typeof(success_data$infant_mortality)
typeof(success_data$median_income)
typeof(success_data$happiness_rating)
```

```{r echo=FALSE, results='hide',message=FALSE,warning=FALSE}
# Success Pct -------------------------------------------------------------

# check over data 
head(success_data)
colnames(success_data)

# create a new tibble for percentages 
success_pct <- tibble("country_name" = success_data$country_name, 
                      "region" = success_data$region)

# add in the economic freedom data
head(success_data$econ_freedom)
success_pct$econ_freedom <- success_data$econ_freedom / 100

# add in the political freedom data 
head(success_data$poli_freedom)
success_pct$poli_freedom <- success_data$poli_freedom / 100

# add in the gdp data 
# since gdp is not easily converted in a pct, will find the max gdp and 
# divide the rest by the max to create a pct 
head(success_data$gdp)
max_gdp <- max(success_data$gdp, na.rm = TRUE)
success_pct$gdp <- success_data$gdp / max_gdp

# add in the unemployment data 
# since high unemployment harms a country's success, we want to subtract 
# the pct by 1 to find a positive idicator (i.e. the employment rate)
head(success_data$unemployment)
success_pct$unemployment <- 1 - (success_data$unemployment / 100)

# add in the crime index data
# since high crime index harms a country's success, we want to subtract 
# the pct by 1 to find a positive idicator (i.e. the safety index)
head(success_data$crime_index)
success_pct$crime_index <- 1 - (success_data$crime_index / 100)

# add in the homicide rate data 
# subtract by 1 to find a positive indicator 
head(success_data$homicide)
success_pct$homicide <- 1 - (success_data$homicide / 100)

# add in the enrollment data 
head(success_data$enrollment)
max_enrollment <- max(success_data$enrollment, na.rm = TRUE)
success_pct$enrollment <- success_data$enrollment / max_enrollment

# add in the literacy data 
head(success_data$literacy)
success_pct$literacy <- success_data$literacy / 100

# add in the life expectancy data
head(success_data$life_expectancy)
max_life <- max(success_data$life_expectancy, na.rm = TRUE)
success_pct$life_expectancy <- success_data$life_expectancy / max_life

# add in the infant mortality data 
# subtract by 1 to find a positive indicator 
head(success_data$infant_mortality)
max_infant <- max(success_data$infant_mortality, na.rm = TRUE)
success_pct$infant_mortality <- 
  1 - (success_data$infant_mortality / max_infant)

# add in the median income data 
head(success_data$median_income)
max_income <- max(success_data$median_income, na.rm = TRUE)
success_pct$median_income <- success_data$median_income / max_income

# add in the happiness rating data 
head(success_data$happiness_rating)
max_happiness <- max(success_data$happiness_rating, na.rm = TRUE)
success_pct$happiness <- success_data$happiness_rating / max_happiness

# calculate the success rating for each country 
# since many countries have at least one NA for their data, we cannot 
# simply add the columns up and will instead need to find the mean
ncol(success_pct)
colnames(success_pct)

success_pct$success <- NA
success_pct$success <- rowMeans(success_pct[,5:14], na.rm=TRUE)

# Success NA --------------------------------------------------------------
# considering success on the basis of five categories: 
# 1. economic prosperity 
# 2. safety 
# 3. education 
# 4. health 
# 5. lifestyle

# each category was given two different variables 

# any country that is missing data for both variables in any of the 
# categories needs to be removed
success_na <- success_pct

colnames(success_pct)

success_na <- 
  success_na[rowSums(is.na(success_na[c("gdp", "unemployment")])) != 2, ]
success_na <- 
  success_na[rowSums(is.na(success_na[c("crime_index", "homicide")])) != 2, ]
success_na <- 
  success_na[rowSums(is.na(success_na[c("enrollment", "literacy")])) != 2, ]
success_na <- 
  success_na[rowSums(is.na(success_na[c("life_expectancy", "infant_mortality")])) != 2, ]
success_na <- 
  success_na[rowSums(is.na(success_na[c("median_income", "happiness")])) != 2, ]

# count number of remaining countries from the original 195 
nrow(success_na)

# Success Combine ---------------------------------------------------------
# create new tibble where the two variables are combines into one for each 
# of the five categories 
success_combine <- tibble("country_name" = success_na$country_name, 
                          "region" = success_na$region, 
                          "poli_freedom" = success_na$poli_freedom, 
                          "econ_freedom" = success_na$econ_freedom)
colnames(success_na)

# economy 
success_combine$econ <- 
  rowMeans(success_na[,c("gdp", "unemployment")], na.rm=TRUE)

# safety
success_combine$safe <- 
  rowMeans(success_na[,c("crime_index", "homicide")], na.rm=TRUE)

# education 
success_combine$educ <- 
  rowMeans(success_na[,c("enrollment", "literacy")], na.rm=TRUE)

# health 
success_combine$health <- 
  rowMeans(success_na[,c("life_expectancy", "infant_mortality")], na.rm=TRUE)

# lifestyle
success_combine$life <- 
  rowMeans(success_na[,c("median_income", "happiness")], na.rm=TRUE)

# add success calculations back 
success_combine$success <- success_na$success

# find the least successful countries
success_bot3 <- success_combine %>% top_n(-10)

# find the most successful countries
success_top3 <- success_combine %>% top_n(10)


# basic statistics --------------------------------------------------------
mean(success_na$success) 
# 0.6117207

min(success_na$success) 
# 0.3012778

success_na %>% top_n(-1) 
# South Sudan

max(success_na$success) 
# 0.9071793

success_na %>% top_n(1) 
# Tawain 

# find rank ---------------------------------------------------------------

# arrange data by success
success_order <- success_na %>% 
  arrange(desc(success))
nrow(success_order)
success_order$rank <- 1:179

# find the least successful countries
success_bot <- success_na %>% top_n(-10)

# find the most successful countries
success_top <- success_na %>% top_n(10)

# create tibbles keeping only countries and success scores
success_bot2 <- tibble("country" = success_bot$country_name, 
                       "success" = success_bot$success)
success_top2 <- tibble("country" = success_top$country_name, 
                       "success" = success_top$success)

# create tibbles keeping only countries and freedom/success scores
success_bot3 <- tibble("country" = success_bot$country_name, 
                       "econ" = success_bot$econ_freedom, 
                       "poli" = success_bot$poli_freedom,
                       "success" = success_bot$success)
success_top3 <- tibble("country" = success_top$country_name, 
                       "econ" = success_top$econ_freedom, 
                       "poli" = success_top$poli_freedom, 
                       "success" = success_top$success)
```

# Research Question 

The objective of the following paper is to track and analyze the effect that economic and political freedom have on the overall success of a nation. 

The paper will first detail which nations are ranked highest/lowest in terms of success as well as which regions throughout the world have the highest rates of success. Further, the paper will investigate if there is a positive correlation between economic freedom and success as well as political freedom and succes, i.e. will the rate of success rise in a consistent, positive trend as the rate of economic freedom or political freedom increases. Finally, the research will track whether there is a greater correlation between the type of government or the economic system in how successful a nation is.

# Background

For the research, two data sets 

# Defining Success 

In order to calculate the success of a nation, the term "success" must first be defined. For the sake of the following research, a successful nation has been defined as one that has attained high levels of economic prosperity, education, safety, and health among its population as well as an overall positive quality of life. When combined, these five categories should work together to create a successful citizenry and nation. 

In order to measure the categories, then, two variables have been assigned to each of the five: economic prosperity (GDP, unemployment rate), safety (crime index, homicide rate), education (secondary school enrollment rate, literacy rate), health (life expectancy, infant mortality), and quality of life/lifestyle (median income, world happiness rating). By measuring each of the five predictors with the same number of variables, no category will be weighted as more crucial in the success of a nation than any of the others. 

By including five different categories and weighting them evenly, the risk of preferential treatment towards one category will be removed. In many previous studies of the success of countries, research focused solely on economic factors (or weighted them heavier than other variables). However, under this current study, the goal is to gain a more robust picture of what success means and looks like for a nation; thus, it becomes necessary to weight economic prosperity the same as the remaining four categories. 

# Calculating Success 

To calculate success, proportions were calculated for each of the ten variables. 

```{r}
head(success_na)
```

# Comparing Economic/Political Freedom and Success 

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# loading country code data 
library(readxl)
country_code <- read_excel("~/Downloads/final_proj/raw_data/country_code.xlsx")

library(tidyverse)
country_code <- as_tibble(country_code)
colnames(country_code)

# clean up data
country_code <- str_split_fixed(country_code$code, " ", 2)
country_code <- as.tibble(country_code)
colnames(country_code)
country_code <- country_code %>% 
  rename(code = `V1`, 
         country = `V2`)
colnames(country_code)

# trim the leading white spaces off of the columns
head(infant2)
country_code$country <- trimws(country_code$country)
country_code$code <- trimws(country_code$code)

# merge country_code with countries from study 
my_countries <- tibble("country" = success_na$country_name)
my_countries$country <- trimws(my_countries$country)

my_countries <- merge(my_countries, country_code, all.x = TRUE)

# merge my_countries with success data
success_code <- success_na
success_code <- success_code %>% 
  rename(country = country_name)

success_code <- merge(success_code, my_countries, all.x = TRUE)
```

## Success Ratings 

The following map depicts the success within each given nation, with the darker shades showing the more successful free nations and the lighter showing the less free:
```{r echo=FALSE}
# create map --------------------------------------------------------------
library(rworldmap)

# countries data frame 
theCountries <- success_code$code

# data frame with success and country code
sDF <- data.frame(country = success_code$code, 
                    success = success_code$success)

# join data frame to map
sMap <- joinCountryData2Map(sDF, joinCode = "ISO3",
                              nameJoinColumn = "country")

# create map
mapCountryData(sMap, nameColumnToPlot="success", catMethod = "quantiles",
               missingCountryCol = gray(.8), mapTitle = "Success of Nations")
```

## Political Freedom

The following map depicts the political freedom within each given nation, with the darker shades showing the more politically free nations and the lighter showing the less free:
```{r echo=FALSE}
# repeat for poli freedom  ------------------------------------------------

# data frame with success and country code
pDF <- data.frame(country = success_code$code, 
                  success = success_code$poli_freedom)

# join data frame to map
pMap <- joinCountryData2Map(pDF, joinCode = "ISO3",
                            nameJoinColumn = "country")

# create map
mapCountryData(pMap, nameColumnToPlot="success", catMethod = "quantiles",
               missingCountryCol = gray(.8), mapTitle = "Poli Freedom of Nations")
```

## Economic Freedom

The following map depicts the economic freedom within each given nation, with the darker shades showing the more economically free nations and the lighter showing the less free:
```{r echo=FALSE}
# repeat for econ freedom -------------------------------------------------

# data frame with success and country code
eDF <- data.frame(country = success_code$code, 
                  success = success_code$econ_freedom)

# join data frame to map
eMap <- joinCountryData2Map(eDF, joinCode = "ISO3",
                            nameJoinColumn = "country")

# create map
mapCountryData(eMap, nameColumnToPlot="success", catMethod = "quantiles",
               missingCountryCol = gray(.8), mapTitle = "Econ Freedom of Nations")
```

## Comparing the Three 

When comparing the three maps to one another, it becomes clear that the three models rate the regions of the world very differently. Success favors rates the Asia-Pacific region more favorably then either political freedom or economic freedom does. Additionally, success's rating of South America falls much more in line with the political freedom model than economic freedom. Lastly, both political and economic freedom rate the southern region of Africa more favorably than the success rate model does. 


# Success by Region 

The following section investigates this topic further by tracking the numerical differences in success between the regions in the three models. 

## Political Freedom 

Under the political freedom model, the most successful region is Europe with an average freedom rating of 0.812 out of 1.0 and the least successful region is the Middle East / North Africa with an average freedom rating of 0.288. The following tibble includes the mean political freedom rating for each of the five regions. 

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# region data 
colnames(success_na)
unique(success_na$region)

# calculate means for each region
mean_ap <- 
  mean(success_na$poli_freedom[success_na$region == "Asia-Pacific"], na.rm = TRUE)
mean_eu <-
  mean(success_na$poli_freedom[success_na$region == "Europe"], na.rm = TRUE)
mean_mena <-
  mean(success_na$poli_freedom[success_na$region == "Middle East / North Africa"], na.rm = TRUE)
mean_ssa <-
  mean(success_na$poli_freedom[success_na$region == "Sub-Saharan Africa"], na.rm = TRUE)
mean_am <-
  mean(success_na$poli_freedom[success_na$region == "Americas"], na.rm = TRUE)

# create a tibble of values
df_region_poli <- tibble("region" = c("Asia-Pacific", "Europe", 
                                 "Middle East / North Africa", 
                                 "Sub-Saharan Africa", "Americas"), 
                    "mean" = c(mean_ap, mean_eu, mean_mena, mean_ssa, mean_am))
```

```{r}
df_region_poli
```

## Economic Freedom 

Under the economic freedom model, the most successful region is Europe with an average freedom rating of 0.689 out of 1.0 and the least successful region is Sub-Saharan Africa with an average rating of 0.546. As is evidenced by the following tibble that includes the mean economic freedom ratings for each of the five regions, there is much less variation between the regions than there is under the other two models:

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# region data 
colnames(success_na)
unique(success_na$region)

# calculate means for each region
mean_ap <- 
  mean(success_na$econ_freedom[success_na$region == "Asia-Pacific"], na.rm = TRUE)
mean_eu <-
  mean(success_na$econ_freedom[success_na$region == "Europe"], na.rm = TRUE)
mean_mena <-
  mean(success_na$econ_freedom[success_na$region == "Middle East / North Africa"], na.rm = TRUE)
mean_ssa <-
  mean(success_na$econ_freedom[success_na$region == "Sub-Saharan Africa"], na.rm = TRUE)
mean_am <-
  mean(success_na$econ_freedom[success_na$region == "Americas"], na.rm = TRUE)
# create a tibble of values
df_region_econ <- tibble("region" = c("Asia-Pacific", "Europe", 
                                 "Middle East / North Africa", 
                                 "Sub-Saharan Africa", "Americas"), 
                    "mean" = c(mean_ap, mean_eu, mean_mena, mean_ssa, mean_am))
```

```{r}
df_region_econ
```

## Success Rating 

Under the success model, the most successful region is, again, Europe with an average rating of 0.711 out of 1.0 and the least successful region is Sub-Saharan Africa with an average rating of 0.491. The following tibble includes the mean success ratings for each of the five regions: 

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# region data 
colnames(success_na)
unique(success_na$region)

# calculate means for each region
mean_ap <- 
  mean(success_na$success[success_na$region == "Asia-Pacific"], na.rm = TRUE)
mean_eu <-
  mean(success_na$success[success_na$region == "Europe"], na.rm = TRUE)
mean_mena <-
  mean(success_na$success[success_na$region == "Middle East / North Africa"], na.rm = TRUE)
mean_ssa <-
  mean(success_na$success[success_na$region == "Sub-Saharan Africa"], na.rm = TRUE)
mean_am <-
  mean(success_na$success[success_na$region == "Americas"], na.rm = TRUE)

# create a tibble of values
df_region <- tibble("region" = c("Asia-Pacific", "Europe", 
                               "Middle East / North Africa", 
                               "Sub-Saharan Africa", "Americas"), 
                  "mean" = c(mean_ap, mean_eu, mean_mena, mean_ssa, mean_am))
```

```{r}
df_region
```

## Comparing the Three 

Under all three models, Europe is consistently the region with the highest average score - making it the most politically/economically free as well as the most successful. There is somewhat more variation in the lowest average scores. Sub-Saharan Africa is the least economically free and least successful while the Middle East / North Africa is the least politically free.  

# Most Successful Nations 



# Least Successful Nations 


# Political Freedom and Success

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# calculate means of proportions ------------------------------------------
success_calc <- success_order

# find success rates for each proportions
suc0 <- success_calc$poli_freedom[success_calc$success == 0]
suc1 <- success_calc$poli_freedom[success_calc$success < .1]
suc2 <- success_calc$poli_freedom[success_calc$success < .2]
suc3 <- success_calc$poli_freedom[success_calc$success < .3]
suc4 <- success_calc$poli_freedom[success_calc$success < .4]
suc5 <- success_calc$poli_freedom[(success_calc$success > .5)&(success_calc$success< .6)]
suc6 <- success_calc$poli_freedom[(success_calc$success > .6)&(success_calc$success< .7)]
suc7 <- success_calc$poli_freedom[(success_calc$success > .7)&(success_calc$success< .8)]
suc8 <- success_calc$poli_freedom[(success_calc$success > .8)&(success_calc$success< .9)]
suc9 <- success_calc$poli_freedom[(success_calc$success > .9)&(success_calc$success< 1.0)]

# calculate means
calc_mean <- c(mean(suc0), mean(suc1), mean(suc2), mean(suc3), mean(suc4),
               mean(suc5), mean(suc6), mean(suc7), mean(suc8), 
               mean(suc9)) 
calc_mean

# create tibbles with proportions
mean_poli_prop <- 
  tibble("prop_success" = c("0 - 0.1", "0.1 - 0.2", "0.2 - 0.3", 
                            "0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6",
                            "0.6 - 0.7", "0.7 - 0.8", "0.8 - 0.9",
                            "0.9 - 1.0"),
         "mean_success" = calc_mean)
```

```{r}
mean_poli_prop
```

# Economic Freedom and Success

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
success_calc <- success_order

# find success rates for each proportions
xsuc1 <- success_calc$econ_freedom[(success_calc$success > .7)&(success_calc$success< .75)]
xsuc2 <- success_calc$econ_freedom[(success_calc$success > .75)&(success_calc$success< .8)]
xsuc3 <- success_calc$econ_freedom[(success_calc$success > .8)&(success_calc$success< .85)]
xsuc4 <- success_calc$econ_freedom[(success_calc$success > .85)&(success_calc$success< .9)]
xsuc5 <- success_calc$econ_freedom[(success_calc$success > .9)&(success_calc$success< .95)]
xsuc6 <- success_calc$econ_freedom[(success_calc$success > .95)&(success_calc$success< 1.0)]

# calculate means
calc_mean <- c(mean(xsuc1), mean(xsuc2), mean(xsuc3), 
               mean(xsuc4),mean(xsuc5), mean(xsuc6)) 
calc_mean

# create tibbles with proportions
mean_econ_prop <- 
  tibble("prop_success" = c("0.70 - 0.75", "0.75 - 0.80", "0.80 - 0.85", 
                            "0.85 - 0.90", "0.90 - 0.95", "0.95 - 1.0"),
         "mean_success" = calc_mean)
```

```{r}
mean_econ_prop
```

# Overall Correlation between Freedom Type and Success

In order to calculate the correlation between economic freedom, political freedom, and success within the same model, a multiple linear regression test will be utilized. 

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
lm_both <- lm(success ~ econ_freedom + poli_freedom, data = success_na)
```

```{r}
lm_both
```

As the outcome of the linear regression shows, economic freedom has a much stronger correlation to the success of a nation than political freedom does. An economic freedom rating of 100 percent will increase a nation's success rating by 62.008 percent whereas a political freedom rating of 100 percent will only increase a nation's success rating by 4.259 percent. 

# Determing the Effect of the Economic Prosperity Indicators 

When calculating the economic freedom of the country, the Heritage foundation utilized both the GDP and unemployments rates in their calculations. Since both of these variables were utilized in the calculation of success as indicators of economic prosperity, it is important to run a second linear regression to test the correlation between success and freedom rates with these two variables removed. 

The second multiple linear regression considers only the remaining four categories - safety, education, health, and lifestyle - when calculating the success of a given country:

```{r echo=FALSE, results='hide',message=FALSE, warning=FALSE}
# calculating success without economic indicators 

# in order to ensure that success was not biased towards economic freedom 
# as gdp and unemployment were both in the calculations of economic freedom 
# and the success variables, need to calculate success removing the economic 
# category to see if it has an effect on the linear regression 

# check over data 
head(success_data)
colnames(success_data)

# create a new tibble for percentages 
success_four <- tibble("country_name" = success_data$country_name, 
                      "region" = success_data$region)

# add in the economic freedom data
head(success_data$econ_freedom)
success_four$econ_freedom <- success_data$econ_freedom / 100

# add in the political freedom data 
head(success_data$poli_freedom)
success_four$poli_freedom <- success_data$poli_freedom / 100

# add in the crime index data
# since high crime index harms a country's success, we want to subtract 
# the pct by 1 to find a positive idicator (i.e. the safety index)
head(success_data$crime_index)
success_four$crime_index <- 1 - (success_data$crime_index / 100)

# add in the homicide rate data 
# subtract by 1 to find a positive indicator 
head(success_data$homicide)
success_four$homicide <- 1 - (success_data$homicide / 100)

# add in the enrollment data 
head(success_data$enrollment)
max_enrollment <- max(success_data$enrollment, na.rm = TRUE)
success_four$enrollment <- success_data$enrollment / max_enrollment

# add in the literacy data 
head(success_data$literacy)
success_four$literacy <- success_data$literacy / 100

# add in the life expectancy data
head(success_data$life_expectancy)
max_life <- max(success_data$life_expectancy, na.rm = TRUE)
success_four$life_expectancy <- success_data$life_expectancy / max_life

# add in the infant mortality data 
# subtract by 1 to find a positive indicator 
head(success_data$infant_mortality)
max_infant <- max(success_data$infant_mortality, na.rm = TRUE)
success_four$infant_mortality <- 
  1 - (success_data$infant_mortality / max_infant)

# add in the median income data 
head(success_data$median_income)
max_income <- max(success_data$median_income, na.rm = TRUE)
success_four$median_income <- success_data$median_income / max_income

# add in the happiness rating data 
head(success_data$happiness_rating)
max_happiness <- max(success_data$happiness_rating, na.rm = TRUE)
success_four$happiness <- success_data$happiness_rating / max_happiness

# calculate the success rating for each country 
# since many countries have at least one NA for their data, we cannot 
# simply add the columns up and will instead need to find the mean
ncol(success_four)
colnames(success_four)

success_four$success <- NA
success_four$success <- rowMeans(success_four[,5:12], na.rm=TRUE)

# remove rows that have NA for both variables in one category 

# any country that is missing data for both variables in any of the 
# categories needs to be removed
success_four <- success_pct

colnames(success_pct)

success_four <- 
  success_four[rowSums(is.na(success_four[c("crime_index", "homicide")])) != 2, ]
success_four <- 
  success_four[rowSums(is.na(success_four[c("enrollment", "literacy")])) != 2, ]
success_four <- 
  success_four[rowSums(is.na(success_four[c("life_expectancy", "infant_mortality")])) != 2, ]
success_four <- 
  success_four[rowSums(is.na(success_four[c("median_income", "happiness")])) != 2, ]

# count number of remaining countries from the original 195 
nrow(success_four)

# run a new linear regression with only the four categories factored into 
# the success score 

lm_both_four <- lm(success ~ econ_freedom + poli_freedom, data = success_four)

lm_five <- lm(success ~ econ + safe + educ + health + life, 
              data = success_combine)
```

```{r}
lm_both_four
```

Given that the correlation coefficients for economic freedom and political freedom remain the same (0.62008 and 0.04259, respectively) even when the economic prosperity category has been removed from consideration, we have ensured that the correlation was not skewed in favor of economic freedom by sharing two of the same indicators.  

When testing the correlation between the five indicators and the success rate for a nation, the results show that health indicators had the strongest correlation with a nation's success while economic factors had the weakest correlation. Thus, it is less surprising that the removal of the economic prosperty category had no effect on the overall correlation between economic freedom/political freedom and success. 

```{r}
lm_five
```
